An efficient acyclic contact planner for multiped robots

### 2016-01-18 body/abstract.tex
nico + glossary

We address the issue of automatically planning discrete sequences of contacts
that characterize a multiped motion in an arbitrary cluttered environment.
To overcome the inherent combinatorial difficulty of this problem, we separate it
in two sub-problems: first, planning a guide path for the root of the robot and
then, generating relevant contacts along this path.
We propose theoretical contributions for each sub-problem, and implement them
as a complete contact planner. 
First, we propose a geometrical characterization of a guide path, the ``reachability condition", which guarantees that a
guide can be mapped into the contact manifold of the robot. With this condition, asserting the``true feasibility" 
of a guide path boils down to performing collision tests, thus avoiding the expensive explicit computation of balanced contact configurations, as opposed to previous approaches.
Then, we propose a sequential algorithm to extend a truly feasible path into a sequence of contacts. For each limb, an off-line sampling approach,
independent from the environment, is run once and for all, and used at run time to provide efficiently candidate configurations for contact.
The best candidate is chosen according to user-defined heuristics. We propose two such heuristics, used in all our experiments, and 
designed for generic robotic applications: a manipulability-based heuristic, EFORT, and a robustness criterion regarding the balance of the robot.
These contributions are implemented within the open-source software Hpp. Our framework allows us to generate complex multi-contact sequences for a large
variety of robots, including HRP-2, HyQ, and a dexterous hand, with interactive performances.

>>>>

We address the issue of automatically planning discrete sequences of contacts
that characterize a multiped motion in an arbitrary cluttered environment.
To overcome the inherent combinatorial difficulty of this problem, we separate it
in two sub-problems: first, planning a guide path for the root of the robot and
then, generating relevant contacts along this path.
We propose theoretical contributions for each sub-problem, and implement them
as a complete contact planner. 
First, we propose a geometrical characterization of a guide path, the ``reachability condition", which guarantees that a
guide can be mapped into the contact manifold of the robot. With this condition, asserting the feasibility
of a guide path boils down to performing collision tests, thus avoiding the expensive explicit computation of balanced contact configurations, as opposed to previous approaches.
Then, we propose a sequential algorithm to extend a feasible guide path into a sequence of contacts. For each limb, an off-line sampling approach,
independent from the environment, is run once and for all, and used at run time to provide efficiently candidate configurations for contact.
The best candidate is chosen according to user-defined heuristics. We propose two such heuristics, used in all our experiments, and 
designed for generic robotic applications: a manipulability-based heuristic, EFORT, and a robustness criterion regarding the equilibrium of the robot.
These contributions are implemented within the open-source software Hpp. Our framework allows us to generate complex multi-contact sequences for a large
variety of robots, including HRP-2, HyQ, and a dexterous hand, with interactive performances.

### 2016-01-18 body/appendix_roms.tex
nico + glossary

\subsection{Step 2: computing the trunk scaling workspace $W^0$ of the robot}
To define the volume $W^0$ of HRP-2, we proceed in an empirical manner.
First, we compute the bounding boxes of the robot trunk, head, and upper legs (Figure~\ref{fig:hrp2_w} -- red shapes).
Then, we perform a scaling of these boxes by a factor $s$. For different values of $s$, we verify the geometrical feasibility condition:
For a large number of configurations that verify the reachability condition, we try to generate contact configurations. We then
empirically select the lowest value of $s$ that provides a reasonnable success rate in all scenarios.
For HRP-2, we choose $s=1.2$. This leads to a success rate higher than 86 \% for all scenarios.

>>>>

\subsection{Step 2: computing the trunk scaling workspace $W^0$ of the robot}
To define the volume $W^0$ of HRP-2, we proceed in an empirical manner.
First, we compute the bounding boxes of the robot trunk, head, and upper legs (Figure~\ref{fig:hrp2_w} -- red shapes).
Then, we perform a scaling of these boxes by a factor $s$. For different values of $s$, we verify the geometrical feasibility condition:
For a large number of configurations that verify the \gls{reachc}, we try to generate contact configurations. We then
empirically select the lowest value of $s$ that provides a reasonnable success rate in all scenarios.
For HRP-2, we choose $s=1.2$. This leads to a success rate higher than 86 \% for all scenarios.

### 2016-01-18 body/conclusion.tex
nico + glossary

Aside from the theoretical contributions, our results demonstrate that our method allows a very interesting compromise between three 
criteria that are hard to conciliate: generality, performance, and quality of the solution, making it the first acyclic contact
planner compatible with interactive applications.
%
\textbf{Regarding generality}, the reachability condition, coupled with an approach based on limb decomposition, 
allows the method to address arbitrary multiped robots. The only pre-requisite is the specification 
of the volumes $W^0$.
%
\textbf{Regarding performance}, our framework is really efficient in addressing both \Pa and \Pb. This results in computation costs close to real-time in
known static environments.
%
\textbf{Regarding the quality of the paths}, a parametrization of the reachability condition allows us to compute
relevant paths in all the presented scenarios, with low rejection rates.
As for \cite{Bouyarmane2009}, failures can still occur, due to the compromise criterion used in computing the guide path.
The low computational burden of our framework however allows for fast re-planning in case of failure.

>>>>

Aside from the theoretical contributions, our results demonstrate that our method allows a very interesting compromise between three 
criteria that are hard to conciliate: generality, performance, and quality of the solution, making it the first acyclic contact
planner compatible with \gls{interactive} applications.
%
\textbf{Regarding generality}, the \gls{reachc}, coupled with an approach based on limb decomposition, 
allows the method to address arbitrary multiped robots. The only pre-requisite is the specification 
of the volumes $W^0$.
%
\textbf{Regarding performance}, our framework is really efficient in addressing both \Pa and \Pb. This results in computation costs close to real-time in
known static environments.
%
\textbf{Regarding the quality of the paths}, a parametrization of the \gls{reachc} allows us to compute
relevant paths in all the presented scenarios, with low rejection rates.
As for \cite{Bouyarmane2009}, failures can still occur, due to the compromise criterion used in computing the guide path.
The low computational burden of our framework however allows for fast re-planning in case of failure.

### 2016-01-18 body/contact.tex
nico + glossary

% !TEX root =  ../main.tex
\section{From a guide path to a discrete sequence of contact configurations ($\mathcal{P}_2$)}
\label{sec:contact}
As an input of this stage, we consider a \textit{geometrically feasible} guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow SE(3)$ for the root of the robot $R$. We now consider the second problem of computing a discrete sequence of contact configurations $\mathbf{Q}^{\overline{0}}$ for the limbs of the robot along $\mathbf{q}^0(t)$.

In this Section we first describe a single contact-generation process, that is how to generate a contact configuration for a given limb 
and root location.
Then, on this method we build an iterative algorithm to generate a discrete sequence of contact configurations in equilibrium.

In what follows a \textit{valid} contact configuration is understood as being collision free, satisfying joint limits, and allowing to maintain the static equilibrium
of the robot. The first two tests are trivial; our criterion to assert efficiently the static equilibrium of the system
is described in Section~\ref{sec:heuristics}.

\subsection{Definition of a contact generator}
\label{sec:single_contact}
We consider a limb $R^k$  and a root configuration $\mathbf{q}^0$.
From \textit{geometrical feasibility} we know that there exists an infinite list of mappings $P = [\pi_0,..,\pi_i, ...]$ that verify \eqref{eq:pi} for $\mathbf{q}^0$.
The issue is then to evaluate locally at least one element of $P$ at $\mathbf{q}^0$.

While exhibiting analytically a $\pi_i$ does not seem tractable, we can iteratively evaluate $\pi_i(\mathbf{q}^0)$ as follows:
\begin{enumerate}
\item Generate randomly a collision-free limb configuration;
\item Project the end-effector onto the closest surface with inverse kinematics;
\item If a valid solution is found, stop. Otherwise repeat from step 1.
\end{enumerate}

@@

We first define $C_{Contact}^{\epsilon} \supset C_{Contact}$ as the set of configurations almost in contact. This means that the minimal distance 
between an effector and a surface of the environment is less than a small value $\epsilon$.
We then apply the following steps:
\begin{enumerate}
\item Generate off-line $N$ valid sample limb configurations $\mathbf{q}^k_i,  0 \leq i < N$;
\item Using the end-effector positions $\mathbf{p}(\mathbf{q}^k_i)$ as indices, store each sample in an octree data structure;
\item Figure~\ref{fig:contact_gen}--b and c--: At runtime, when contact creation is required, retrieve from the octree the list of samples $S \subset C_{Contact}^{\epsilon}$ close to contact;
\item According to a user-defined heuristic $h$, sort the elements of $S$;
\item Figure~\ref{fig:contact_gen}--d and e--: Select the first configuration of $S$ (or fail if $S$ is empty). Project the configuration onto contact with inverse kinematics;
\item If a valid solution is found, stop. Otherwise remove the element from $S$ and go back step 5.
\end{enumerate}

This approach has two main advantages.
First, this allows us to select a large number of candidates in a single proximity request.
Having several candidates is interesting, because it allows to compare them using a user-selected heuristic $h$, thus obtaining
a locally-optimal candidate.
Furthermore, the fact that the candidates are already close to contact increases the odds that the inverse kinematics will converge to a valid solution in a small number of iterations.
Regarding completeness, it is immediate to verify that as $N$ grows, the completeness of the method is restored.
$N$ is a parameter that allows us to specify the trade-off between completeness and efficiency.
The reader is invited to refer to our previous work for an extensive discussion on the optimal value of $N$ \citep{Tonneau2014}.

@@

As an input, we consider the guide path $\mathbf{q}^0(t)$, automatically discretized into a sequence of $j$ key placements:  
\begin{equation*}
	\mathbf{Q}^0 = [\mathbf{q}^0_{0}; \mathbf{q}^0_{i}; ..., \mathbf{q}^0_{j-1}]
\end{equation*} 
where $\mathbf{q}^0_{0}$ and $\mathbf{q}^0_{j-1}$ respectively correspond to the start and goal configurations. To ensure continuity in the contact transition phases, we rewrite $\pi$ under the following recursive form for any $0<i<j$:
\begin{equation*}
    \pi\colon\left\{
    \begin{aligned}		
        \mathbf{Q}^0 \in C_{reach} & \longrightarrow C_{contact} \\
        %\mathbf{q}^{0}_0 &  \longrightarrow  \mathbf{q}_{start} \\
        \mathbf{q}^{0}_i &  \longrightarrow  \mathbf{q}^{0}_i \oplus g(\mathbf{q}_{i - 1},\mathbf{q}^{0}_i) 
    \end{aligned}
    \right.
\end{equation*} 
$g$ is the method that extends a root configuration into a full-body configuration. At each step, it tries to generate a contact configuration that preserves as much as possible
the previous contacts while allowing for static equilibrium. The objective is thus to characterize $g$.
We initialize the recurrence with $\pi(\mathbf{q}^0_{0}) = \mathbf{q}_0$ the initial configuration of the robot.

@@

These three rules provide a deterministic manner to generate contacts along the discretized guide path.
For a given truly feasible path, our current implementation does not guarantee that the planner will succeed in generating a configuration in static equilibrium, because
of the deterministic approach used to break the combinatorial.
However, in practice the planner is successful in the large majority of cases, as discussed in Section~\ref{sec:perf}, thanks to 
the heuristics that we use for contact selection.

>>>>

% !TEX root =  ../main.tex
\section{From a guide path to a discrete sequence of contact configurations ($\mathcal{P}_2$)}
\label{sec:contact}
Our planner computes guide paths in \gls{creach0}, an over-approximation of the set of \gls{equilibrium feasible} paths, \gls{ccequil0}.
As an input of this stage, we however assume an \gls{equilibrium feasible} guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow \mathbb{R}^r$ for the root of the robot $R$.
If this is not the case, our planner will fail rapidly, thus allowing replanning, as discussed in Section TODO.
We now consider the second problem of computing a discrete sequence of contact configurations $\mathbf{Q}^{\overline{0}}$ for the limbs of the robot along $\mathbf{q}^0(t)$.

In this Section we first describe a single contact-generation process, that is how to generate a contact configuration for a given limb 
and root location.
Then, we propose an iterative algorithm to generate a discrete sequence of contact configurations in static equilibrium.

Our criterion to assert efficiently the static equilibrium of the system
is described in Section~\ref{sec:heuristics}.

\subsection{Definition of a contact generator}
\label{sec:single_contact}
We consider a limb $R^k$  and a root configuration $\mathbf{q}^0$.
From \glslink{contact feasible}{\textit{contact feasibility}} we know that there exists at least one $\mathbf{q}^{\overline{0}}$ that verifies $\mathbf{q}^k \oplus \mathbf{q}^{\overline{k}} \in$ \gls{ccontact}.
The issue is then to exhibit locally this $\mathbf{q}^{\overline{0}}$.

While exhibiting analytically a $\mathbf{q}^{\overline{0}}$ does not seem tractable, we can iteratively try to generate one as follows:
\begin{enumerate}
\item Generate randomly a collision-free limb configuration;
\item Project the end-effector onto the closest surface with inverse kinematics;
\item If a valid solution is found, stop. Otherwise repeat from step 1.
\end{enumerate}

@@

We first define $C_{Contact}^{\epsilon} \supset C_{Contact}$ as the set of configurations almost in contact. This means that the minimal distance 
between an effector and a surface of the environment is less than a small value $\epsilon$.
We then apply the following steps:
\begin{enumerate}
\item Generate off-line $N$ valid sample limb configurations $\mathbf{q}^k_i,  0 \leq i < N$;
\item Using the end-effector positions $\mathbf{p}(\mathbf{q}^k_i)$ as indices, store each sample in an octree data structure;
\item At runtime, when contact creation is required, retrieve from the octree the list of samples $S \subset C_{Contact}^{\epsilon}$ close to contact (Figure~\ref{fig:contact_gen} (b) and (c));
\item According to a user-defined heuristic $h$, sort the elements of $S$;
\item Select the first configuration of $S$. Project the configuration onto contact with inverse kinematics. If $S$ is empty, stop (failure case) (Figure~\ref{fig:contact_gen} (d) and (e));
\item If a valid solution is found, stop (success case). Otherwise remove the element from $S$ and go back step 5.
\end{enumerate}

This approach has two main advantages.
First, this allows us to select a large number of candidates in a single proximity request.
Having several candidates is interesting, because it allows to compare them using a user-selected heuristic $h$, thus obtaining
a locally-optimal candidate.
Furthermore, the fact that the candidates are already close to contact increases the odds that the inverse kinematics will converge to a valid solution in a small number of iterations.
Regarding convergence, it is immediate to verify that as $N$ grows, the probability of finding a solution if it exists converges to 1.
$N$ is a parameter that allows us to specify the trade-off between exhaustiveness and efficiency.
The reader is invited to refer to our previous work for an extensive discussion on the optimal value of $N$ \citep{Tonneau2014}.

It should be noted that a failure case for our contact generator does not necessarily mean that the planner will globally fail. Indeed,
for a given root configuration, if contact configurations are found for other limbs, resulting in static equilibrium, the configuration will be accepted despite
the failure to generate a contact.

@@

As an input, we consider the guide path $\mathbf{q}^0(t)$, automatically discretized into a sequence of $j$ key configurations:  
\begin{equation*}
	\mathbf{Q}^0 = [\mathbf{q}^0_{0}; \mathbf{q}^0_{i}; ..., \mathbf{q}^0_{j-1}]
\end{equation*} 
where $\mathbf{q}^0_{0}$ and $\mathbf{q}^0_{j-1}$ respectively correspond to the start and goal configurations. To ensure continuity in the contact transition phases, we propose a recursive mapping $\pi$, for any $0<i<j$:
\begin{equation*}
    \pi\colon\left\{
    \begin{aligned}		
        \mathbf{Q}^0 \in C_{equil}^0 & \longrightarrow C_{equil} \\
        %\mathbf{q}^{0}_0 &  \longrightarrow  \mathbf{q}_{start} \\
        \mathbf{q}^{0}_i &  \longrightarrow  \mathbf{q}^{0}_i \oplus g(\mathbf{q}_{i - 1},\mathbf{q}^{0}_i) 
    \end{aligned}
    \right.
\end{equation*} 
$g$ is the method that extends a root configuration into a full-body configuration. At each step, it tries to generate a contact configuration that preserves as much as possible
the previous contacts while allowing for static equilibrium. The objective is thus to characterize $g$.
We initialize the recurrence with $\pi(\mathbf{q}^0_{0}) = \mathbf{q}_0$ the initial configuration of the robot.

@@

These three rules provide a deterministic manner to generate contacts along the discretized guide path.
For a given \gls{equilibrium feasible} path, our current implementation does not guarantee that the planner will succeed in generating a configuration in static equilibrium, because
of the deterministic approach used to break the combinatorial.
However, in practice the planner is successful in the large majority of cases, as discussed in Section~\ref{sec:perf}, thanks to 
the heuristics that we use for contact selection.

### 2016-01-18 body/equilibrium.tex
nico + glossary

\subsubsection{Conditions for static balance.}
We first define the variables of the problem, for a robot $R$ and $e$ contact points, expressed in world coordinates:
\begin{itemize}
\item $\mathbf{c} \in \mathbb{R}^3$ is the robot center of mass (COM);
\item $\mathbf{L}  \in \mathbb{R}^3 $ is the angular momentum at the COM;
\item $\mathbf{m} \in \mathbb{R}$ is the robot mass;
\item $\mathbf{g} = [0,0,-9.81]^T$ is the gravity acceleration;
\item $\mu$ is the friction coefficient;
\item for the i-th contact point $1 \leq i \leq e$:
	\begin{itemize}
	\item $\mathbf{p_i}$ is the contact position;
	\item $\mathbf{f_i}$ is the force applied at $\mathbf{p_i}$;
	\item $\mathbf{n}_i,\mathbf{t}_{i1},\mathbf{t}_{i2}$ form a local Cartesian coordinate system centered at $\mathbf{p_i}$. $\mathbf{n}_i$ is aligned
	with the contact surface normal, and the $\mathbf{t}_i$s are tangent vectors.
	\end{itemize}
\end{itemize}

>>>>

\subsubsection{Conditions for static equilibrium.}
We first define the variables of the problem, for a robot $R$ and $e$ contact points, expressed in world coordinates:
\begin{itemize}
\item $\mathbf{c} \in \mathbb{R}^3$ is the robot center of mass (COM);
\item $\mathbf{L}  \in \mathbb{R}^3 $ is the angular momentum at the COM;
\item $\mathbf{m} \in \mathbb{R}$ is the robot mass;
\item $\mathbf{g} = [0,0,-9.81]^T$ is the gravity acceleration;
\item $\mu$ is the friction coefficient;
\item for the i-th contact point $1 \leq i \leq e$:
	\begin{itemize}
	\item $\mathbf{p_i}$ is the contact position;
	\item $\mathbf{f_i}$ is the force applied at $\mathbf{p_i}$;
	\item $\mathbf{n}_i,\mathbf{t}_{i1},\mathbf{t}_{i2}$ form a local Cartesian coordinate system centered at $\mathbf{p_i}$. $\mathbf{n}_i$ is aligned
	with the contact surface normal, and the $\mathbf{t}_i$s are tangent vectors.
	\end{itemize}
\end{itemize}
